---
title: "moDiNA Simulation Evaluation — Balanced vs Unbalanced AUC comparison"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(pROC)
library(data.table)
library(ggplot2)
library(patchwork)
library(knitr)
```

This report compares the node-ranking AUC across two families of simulations:

* **Balanced** sample-size simulations (`run_balanced_simulations.py`), where both
  contexts share the same number of samples (n = 100, 300, 500, 1000).
* **Unbalanced** sample-size simulations (`run_unbalanced_simulations.py`), where
  Context 1 is fixed at 300 samples and Context 2 grows with the ratio (1:1 → 1:10).

AUC is computed exactly as in `bin/evaluation_auc.R`: how well a config's node ranking
separates the ground-truth (perturbed) nodes from the rest. The two families are shown
as a stacked two-panel heatmap — **Panel A: Balanced**, **Panel B: Unbalanced**.

```{r config}
BALANCED_BASE   <- "/nfs/data/DyHealthNet/moDiNA_data/balanced"
UNBALANCED_BASE <- "/nfs/data/DyHealthNet/moDiNA_data/unbalanced"

# --- AUC helpers (mirrors bin/evaluation_auc.R) ---
calculate_AUC_per_simulation <- function(ranking, ground_truth, node_ranking = TRUE) {
  ranking <- ranking[order(rank)]
  if (node_ranking) {
    ranking[, is_ground_truth := ifelse(node %in% ground_truth$node, 1, 0)]
  } else {
    ranking[, is_ground_truth := ifelse(edge %in% ground_truth$edge, 1, 0)]
  }
  roc_obj <- roc(ranking$is_ground_truth, ranking$rank, quiet = TRUE)
  return(auc(roc_obj))
}

calculate_AUC_for_row <- function(ground_truth_file, ranking_file, row_num, node_ranking = TRUE) {
  tryCatch({
    ground_truth <- fread(ground_truth_file)
    ranking <- fread(ranking_file)
    return(calculate_AUC_per_simulation(ranking, ground_truth, node_ranking))
  }, error = function(e) {
    cat("\n=== ERROR in row", row_num, "===\n")
    cat("Ground truth file:", ground_truth_file, "\n")
    cat("Ranking file:", ranking_file, "\n")
    cat("Error message:", e$message, "\n")
    cat("========================\n")
    return(NA_real_)
  })
}

# summary.csv stores absolute work/ paths that may be cleaned; the same files are
# also published under each out_dir. Prefer the published copy, fall back to the
# recorded path if the reconstruction is missing.
resolve_path <- function(out_dir, subdir, recorded) {
  cand <- file.path(out_dir, subdir, basename(recorded))
  ifelse(file.exists(cand), cand, recorded)
}

# --- Setting discovery (one helper per run family) ---
# Balanced dirs are named "n<samples>" (equal samples per context).
discover_balanced <- function(base) {
  dirs <- list.dirs(base, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl("/n[0-9]+$", dirs)]
  s <- data.table(out_dir = dirs, label = basename(dirs))
  s[, order_key := as.numeric(sub("n", "", label))]
  s[, display  := as.character(order_key)]  # x-axis tick, e.g. "300"
  s[order(order_key)]
}

# Unbalanced dirs are named "1to<ratio>" (Context1 fixed, Context2 scaled).
discover_unbalanced <- function(base) {
  dirs <- list.dirs(base, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[grepl("/1to[0-9.]+$", dirs)]
  s <- data.table(out_dir = dirs, label = basename(dirs))
  s[, order_key := as.numeric(sub("1to", "", label))]
  s[, display  := sub("1to", "1:", label)]  # x-axis tick, e.g. "1:10"
  s[order(order_key)]
}

settings_bal <- discover_balanced(BALANCED_BASE)
settings_unb <- discover_unbalanced(UNBALANCED_BASE)

kable(settings_bal[, .(label, samples_per_context = order_key, out_dir)],
      caption = "Discovered balanced settings")
kable(settings_unb[, .(label, ratio = order_key, out_dir)],
      caption = "Discovered unbalanced settings")
```

```{r build-results}
# Load every summary.csv for a settings table, score AUC per row, and aggregate.
# Returns one row per (setting × node_metric × edge_metric × algorithm) with an
# ordered `setting` factor (levels by order_key, labels = the family's display ticks).
build_results <- function(settings, run_type) {
  dt_list <- lapply(seq_len(nrow(settings)), function(i) {
    out_dir <- settings$out_dir[i]
    summary_file <- file.path(out_dir, "summary.csv")
    if (!file.exists(summary_file)) {
      warning("Missing summary.csv for ", out_dir)
      return(NULL)
    }
    d <- fread(summary_file)
    d[, ranking_file := mapply(resolve_path, out_dir, "rankings", ranking_file)]
    d[, ground_truth_nodes := mapply(resolve_path, out_dir, "copula_simulation", ground_truth_nodes)]
    d[, auc := mapply(calculate_AUC_for_row, ground_truth_nodes, ranking_file,
                      row_num = .I, SIMPLIFY = TRUE)]
    d[, `:=`(setting = settings$label[i], order_key = settings$order_key[i])]
    d
  })
  dt <- rbindlist(dt_list, use.names = TRUE, fill = TRUE)

  # Normalise empty metric labels (as in evaluation_auc.R)
  dt[node_metric == "", node_metric := "None"]
  dt[edge_metric == "", edge_metric := "None"]

  results <- dt[, .(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc   = sd(auc, na.rm = TRUE),
    n        = .N
  ), by = .(setting, order_key, node_metric, edge_metric, algorithm)]

  # Ordered factors: settings by order_key (shown as the family's display tick),
  # edge metric by appearance.
  lvl <- settings[order(order_key)]
  results[, setting := factor(setting, levels = lvl$label, labels = lvl$display)]
  results[, edge_metric := factor(edge_metric, levels = unique(edge_metric))]
  results[, run_type := run_type]
  setorder(results, order_key, edge_metric)
  results
}

res_bal <- build_results(settings_bal, "Balanced")
res_unb <- build_results(settings_unb, "Unbalanced")
```

## AUC heatmap: Balanced (A) vs Unbalanced (B)

```{r heatmap, fig.width=8, fig.height=8}
make_heatmap <- function(results, x_label) {
  ggplot(results, aes(x = setting, y = edge_metric, fill = mean_auc)) +
    geom_tile(color = "black") +
    geom_text(aes(label = sprintf('%.2f', mean_auc)), size = 3) +
    facet_grid(algorithm ~ ., scales = "free", space = "free") +
    scale_fill_gradient(low = 'white', high = '#a32e28', name = 'Mean AUC',
                        guide = guide_colorbar(barwidth = 1.5, barheight = 10),
                        limits = c(0, 1.0)) +
    labs(x = x_label, y = "Edge Metric") +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      panel.spacing.x = unit(1, 'lines'),
      axis.text.x = element_text(size = 10),
      axis.text.y = element_text(size = 10),
      strip.text.y = element_text(size = 10, angle = 0),
      strip.text.x = element_text(size = 10),
      axis.title.x = element_text(size = 10),
      axis.title.y = element_text(size = 10),
      strip.background = element_rect(fill = 'grey90', color = 'black', linewidth = 0.5)
    )
}

p_bal <- make_heatmap(res_bal, "Samples per context (n)")
p_unb <- make_heatmap(res_unb, "Sample-size ratio (Context1:Context2)")

p_combined <- (p_bal / p_unb) +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect")
print(p_combined)
ggsave("balanced_vs_unbalanced_auc_heatmap.png", plot = p_combined,
       width = 8, height = 8, dpi = 300)
```

## Numeric summary (mean ± sd AUC over simulations)

```{r table}
res_all <- rbindlist(list(res_bal, res_unb), use.names = TRUE, fill = TRUE)
tab <- res_all[, .(run_type, setting, edge_metric, node_metric, algorithm,
                   `mean AUC` = sprintf('%.3f', mean_auc),
                   `sd AUC`   = sprintf('%.3f', sd_auc),
                   n_sim = n)]
kable(tab, caption = "AUC per run type × setting × edge metric")
```
